Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-05-18 MS/ms.Rmd
Added one more important point to existing exclusion criteria. That is, we should only focus on studies that collate data from literature

A critically important feature of meta-analaysis that provides improved precision when estimating overall effects is the ability to account for sampling error - this can be derived assuming the sampling distribution of a given effect size is known [].

Despite the widespread adoption and well established methodological procedures of meta-analysis it is often criticised for mixing 'apples and oranges' - in otherwords, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widspread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., ]. In adidtion, they often inlcude a diversity of study species which can vary in responses.

@@

We excluded plant meta-analyses because they were often focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions) that did not focus on organism physiological function across species. Meta-analyses needed to have collected physiological trait data from the literature including traits, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (heart rates, sprint speed, swim speed, growth rates) and immune traits (cell function, antibodies, ect). Studies were also included if they measured reproductive allocation, survival / mortality or quantified scaling relationships with body size (e.g., metabolic scaling) across taxa. Studies on behavior were only included if they were explicitly interested in proximate physiological drivers of behavioral variation. Meta-analyses quantifying phenology, differences in body sizes between the sexes (i.e., SSD) and those focused on validating physiological methodology were excluded because the underlying question was not physiological in nature.

@@

#### *New Effect Size Metrics that Account for Variability in Treatments*

>>>>

Meta-analysis is a well-developed research field with a long history in the fields of statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis, and is becoming increasingly more prominant the field of comparative physiology. Meta-analysis involves strict literature-based search protocols and data extraction (i.e., following the PRISMA protocol, e.g., [@ODea2021]) to achieve two major goals, 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error []. The inverse sampling error is used to weight higher powered studies in the analysis and make overall esatimtes more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on esatimytes, even if the overall mean itself is unbiased.

Despite the widespread adoption and well established methodological procedures of meta-analysis it is often criticised for mixing 'apples and oranges' - in otherwords, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widspread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. In addition, they often include a diversity of study species which can vary in responses.

@@

We excluded plant meta-analyses because they were often focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions) that did not focus on organism physiological function across species. Meta-analyses needed to have collected physiological trait data from the literature including traits, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (heart rates, sprint speed, swim speed, growth rates) and immune traits (cell function, antibodies, ect). Studies were also included if they measured reproductive allocation, survival / mortality or quantified scaling relationships with body size (e.g., metabolic scaling) across taxa. Studies on behavior were only included if they were explicitly interested in proximate physiological drivers of behavioral variation. Meta-analyses quantifying phenology, differences in body sizes between the sexes (i.e., SSD) and those focused on validating physiological methodology were excluded because the underlying question was not physiological in nature. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature.

@@

A very common occurance in comparative physiology is to manipulate the tempearture organisms experience and measure some physiologcal rate (e.g., metabolic rate). Physiological rates are often compared using $Q_{10}$ values (e.g., [@Rodger2021]), which describe the multiplicative change in physiological rates across a 10&deg;C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to weight effect size estimates by sampling variance as is typical in meta-analysis. Recognition of the similarity between *lnRR* and $Q_{10}$ means that we can derive $Q_{10}$ based effect sizes using the well known mathematical properties of *lnRR*, and improve upon our ability to account for sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017]. As such, here we derive a series of $Q_{10}$ based effect sizes along with their associated sampling variances that allow one to compare both changes in mean and variance of physiological rates.

#### *Comparing changes in mean physiological rates*

Prior to showing how the relevant $Q_{10}$ effect size can be calculated it can be helpful to understand its similarities to *lnRR*. The *lnRR* described by @Hedges1999 and extended by  @Lajeunesse2011 and can be calculated as follows:

$$
\begin{equation} 
  \begin{aligned}
  lnRR = ln\left ( \frac{X_{1}}{X_{2}}\right )
  \end{aligned}
  (\#eq:lnRR)
\end{equation} 
$$

$$
\begin{equation} 
  s_{lnRR} = \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right )
  (\#eq:slnRR)
\end{equation} 
$$
In equation \@ref(eq:lnRR), $X_{1}$ is the mean of group 1, often a control group, where as $X_{2}$ is the mean of group 2 (e.g., a treatment group). The mean of $X_{i}$ for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the measurement variable is ratio scale. Log transformation of this ratio makes the effect size normally distributed. Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where, $SD_{1}^2$ and $SD_{2}^2$ is the variance in group 1 and 2, respectively and $N_{1}$ and $N_{2}$ is the sample size in group 1 and 2.

The equations for *lnRR* and its sampling variance allow us to easily extend this to $Q_{10}$ based effect sizes. Recall that $Q_{10}$ is described as follows:

$$
\begin{equation} 
  Q_{10} = \left( \frac{R_{2}}{R_{1}} \right)^{ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
  (\#eq:q10)
\end{equation} 
$$
Here, $R_{1}$ and  $R_{2}$ are mean physiological rates and  $T_{1}$ and  $T_{2}$ are the temperatures that these rates are measured. Log transformation of equation \@ref(eq:q10) leads to the following log transformed $Q_{10}$:

$$
\begin{equation} 
  lnRR_{Q_{10}} = ln\left( \frac{R_{2}}{R_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10)
\end{equation} 
$$
Equation \@ref(eq:lnq10) is essentially a temperature corrected equivalent to lnRR when the numerator and denomenator are measured at different temperatures. This allows one to compare the mean of two temperature treatments directly regardless of the temperatures that these groups have been measured. Here, we will refer to this as the log $Q_{10}$ response ratio, $lnRR_{Q_{10}}$. Recognition of this equivalence means that we can easily calculate the sampling variance for equation \@ref(eq:lnq10) as follows:

$$
\begin{equation} 
  s_{lnRR_{Q_{10}}} = \left( \frac{SD_{2}^2}{R^2_{2}N_{2}} + \frac{SD_{1}^2}{R^2_{1}N_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
  (\#eq:Vlnq10)
\end{equation} 
$$

#### *New Effect Size Metrics that Compare Changes in Variability Across Treatments*

@Nakagawa2015 recently proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). Like *lnRR*, *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variablity between two groups. We refer readers to @Nakagawa2015 for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling varaince) as follows:

$$
\begin{equation} 
lnVR_{Q_{10}} = ln\left( \frac{SD_{2}}{SD_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10VR)
\end{equation} 
$$
$$
\begin{equation} 
s_{lnVR_{Q_{10}}} = \left( \frac{1}{2\left( N_{2}-1 \right)} + \frac{1}{2\left(N_{1} - 1\right)} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
(\#eq:slnq10VR)
\end{equation} 
$$

Equations \@ref(eq:lnq10VR) and \@ref(eq:slnq10VR) describe the change in physiological rate variance (eqn \@ref(eq:lnq10VR)) across a 10&deg;C temperature change along with its sampling variance (eqn \@ref(eq:slnq10VR)). While this is a useful metric, as discussed by Nakagawa *et al* @nakagawa2015meta there is often a strong mean-variance relationship that needs to be accounted for in analysing changes in variance. As such, we can calculate the coefficient of variation, which standardises changes in variance for changes in means as follows:

$$
\begin{equation} 
lnCVR_{Q_{10}} = ln\left( \frac{\text{CV}_{2}}{\text{CV}_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10CVR)
\end{equation} 
$$
$$
\begin{equation} 
s_{lnCVR_{Q_{10}}} = \left[ \frac{(\text{SD}_{1})^2}{N_{1}({R}_{1})^2} + \frac{(\text{SD}_{2})^2}{N_{2} ({R}_{2})^2} + \frac{1}{2\left( N_{1}-1 \right)} + \frac{1}{2\left(N_{2} - 1\right)} \right]{ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
(\#eq:slnq10CVR)
\end{equation} 
$$

where $CV$ is the coefficient of variation defined as $SD / R$.
